---
title: "Numeric vectors with tags"
output: rmarkdown::html_vignette
author: Matt Kerlogue
date: 2023-05-31
vignette: >
  %\VignetteIndexEntry{Numeric vectors with tags}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(shrthnd)
```

The development of the `{shrthnd}` package is heavily influenced by experience
of working with statistical datasets published by governments and international
bodies, especially departments and agencies in the UK producing outputs as
part of the [UK statistical system](https://www.statisticsauthority.gov.uk).

While data is increasingly released in machine readable formats or through APIs,
there are still a large number of data products that continue to be released in
spreadsheets and historical data from these institutions is often only
available in spreadsheets. Beyond layout issues, such as the use of header and
footer rows to communicate related information, it is not uncommon to encounter
columns in these spreadsheets that contain a mix of numeric and non-numeric
content. This non-numeric content may sometimes be the only content of a cell
(to explain why there is no numeric value) or alongside a numeric value (to
qualify, caveat or otherwise explain something about the value).

The most common approach in data processing when encountering these sorts of
issues is simply to scrub the vector of the non-numeric components and coerced
it into a numeric vector. However, these tags often convey useful information
which you may wish to retain.

# Introducing `shrthnd_num()`

The `shrthnd_num()` data type builds on `vctrs::new_rcrd()` to split numeric
and non-numeric components while keeping them attached to each other. In
practice a `shrthnd_num()` can be thought of as a `numeric()` and `character()`
vector that have been coupled together. Specifically it has a `num` component
representing the numeric value and a `tag` component representing the
non-numeric shorthand, symbol or marker.

Let us create the vector `x` with seven values:

```{r}
x <- c("12", "34.567", "[c]", "NA", "56.78[e]", "78.9", "90.123[e]")
x
```

The first, second and sixth values in this vector are purely numeric (`12`,
`34.567` and `78.9`). The third value is a shorthand symbol (`"[c]"`) denoting
that the value has been suppressed because it is confidential. The fourth value
is a missing value (`"NA"`). The fifth and seventh values contain both numeric
information (`56.78` and `90.123` respectively) but also shorthand (`"[e]"`) to
denote that these values are estimated. Depending on what processing we wish to
do with this vector in the future it might be useful to know that a value has
been suppressed or estimated.

Using `as.numeric()` on this vector will result in all of the values containing
any non-numeric element to be converted to a missing value, causing us to lose
all the information of the third, fifth and seventh values in the vector.

```{r}
as.numeric(x)
```

We could scrub the non-numeric elements of the vector, but we still lose the
information provided by the shorthand.

```{r}
as.numeric(gsub("[^0-9.]", "", c(x)))
```

The `shrthnd_num()` function, however, allows us to retain both sets of
information, and we can easily coerce a `shrthnd_num()` vector into a regular
base R `numeric()` vector. We can also easily access the shorthand or symbol
tags with the `shrthnd_tags()` function.

```{r}
sh_x <- shrthnd_num(x)

sh_x

as.numeric(sh_x)

shrthnd_tags(sh_x)
```

The `shrthnd_list()` function provides a summary of the tags contained in a
`shrthnd_num()` vector, their frequency and positions in the vector.

```{r}
shrthnd_list(sh_x)
```

We saw above how `as.numeric()` converts a `shrthnd_num()` to a numeric vector,
`as.character()` will similarly convert a `shrthnd_num()` to a character vector
as if it were a numeric vector. Instead to print a character vector that
combines the numeric and non-numeric components we can use `as_shrthnd()`.

```{r}
as.character(sh_x)

as_shrthnd(sh_x)
```

# Making a shrthnd_num

You can make a `shrthnd_num()` vector in two ways: using `shrthnd_num()` to
convert a character vector containing numeric and non-numeric components, or
`make_shrthnd_num()` to merge a vector of numbers and a vector of character
strings.

## Conversion to a shrthnd_num

You convert a character vector containing shorthand using `shrthnd_num()`. In
addition to the character vector you can also supply additional arguments to
control the behaviour of the conversion and the resulting display of the vector.

```{r}
shrthnd_num(
  x,
  shorthand = NULL,
  na_values = "NA",
  digits = 2L,
  paren_nums = c("negative", "strip"),
  dec = ".",
  bigmark = ","
)
```

The `shorthand` argument allows you to pass a character vector of shorthand,
symbols and markers that you want to validate against, i.e. you can cause the
conversion to throw an error if it detects shorthand that is not in this vector.

The `na_values` argument is used to determine values that should be ignored
when identifying shorthand tags and converted to missing values when extracting
the numeric component.

The `digits`, `dec` and `bigmark` arguments are passed on to `formatC()` in the
formatting of the numeric component when formatting and printing the vector.

The `paren_nums` argument determines how to handle numbers in parenthesis, i.e.
whether to consider a number in parenthesis as a negative number (as is
commonly used in accounting formats, and the default setting) or whether to
just strip the parenthesis from the number before its conversion.

The coercion to a `numeric()` vector is handled by `utils::type.convert()`.

## Making a shrthnd_num from scratch

You can use `make_shrthnd_num()` to create a `shrthnd_num()` from a `numeric()`
and `character()` vector of the same length.

```{r}
make_shrthnd_num(c(1:3, NA, 4:5, NA), c("", "", "", "[c]", "", "[e]", NA))
```

# Coercion, maths and statistics

Generally a `shrthnd_num()` should behave like a `numeric()` vector. For
example, using `is.na()` will return `TRUE` where the numeric value is missing
and `FALSE` where the numeric value is not missing. Or, if you use `c()` to
combine a `shrthnd_num()` with another vector it will first coerce the vector
to numeric so that R can proceed from there.

```{r}
is.na(sh_x)

c(sh_x, 1)

c(sh_x, "c")
```

However, in keeping with base R practice around complex numeric objects such
as `Date()`, `difftime()` and `POSIXct()`, using `is.numeric()` on a
`shrthnd_num()` vector will return `FALSE`. Use `is_shrthnd_num()` to test
if a vector is a `shrthnd_num()` vector.

```{r}
is.numeric(sh_x)

is_shrthnd_num(sh_x)
```

Through `vctrs::vec_arith()` and `vctrs::vec_math()` there is generalised
support for arithmetic and mathematical operations on a `shrthnd_num()` vector.
Bespoke methods have also been added for some functions which are not directly
supported, such as `median()` and `quantile()`, so that they can easily
work with the numeric components of the `shrthnd_num()` vector.

```{r}
x <- c("12", "34.567", "[c]", "NA", "56.78[e]", "78.9", "90.123[e]")
sh_x <- shrthnd_num(x, c("[c]", "[e]"))

sh_x * 2

2 + sh_x

sum(sh_x, na.rm = TRUE)

range(sh_x, na.rm = TRUE)

mean(sh_x, na.rm = TRUE)
```

# Working with shrthnd tags

The `shrthnd_tags()` function allows us to access the tag components of a
`shrthnd_num()`. It has a related function `shrthnd_unique_tags()` which will
return a unique list of tags, and is simply a convenience function in place of
`unique(shrthnd_tags(x))`.

```{r}
shrthnd_tags(sh_x)

shrthnd_unique_tags(sh_x)
```

The base R functions for value matching work with the numeric component of a
`shrthnd_num()` vector. Separate tag locator functions have been used to
support matching the tag components of a `shrthnd_num()` vector.

`tag_match()` returns an integer vector showing the first location of the tag
provided while `tag_in()` will return `TRUE` or `FALSE` depending on whether
the tag is in the vector's shorthand.

```{r}
tag_match(sh_x, "[e]")

tag_in(sh_x, "[e]")
```

To locate where a specific tag is used in a vector use `where_tag()`, which is
equivalent to computing `tags == tag`. To identify if a value has a tag,
irrespective of its value use `any_tag()`, which is equivalent to
`!is.na(tags)`.

```{r}
where_tag(sh_x, "[e]")

any_tag(sh_x)
```

Using `is.na()` on a `shrthnd_num()` will assess if the numeric component is
missing. To identify if tags are missing use `is_na_tag()`, which is equivalent
to `is.na(tags)`. To identify if both the numeric and tag component is missing
use `is_na_both()`.

```{r}
is_na_tag(sh_x)

is_na_both(sh_x)
```

Finally, you can locate the positions of a specific tag, tagged values or
untagged values using a set of `locate_*()` functions, which are convenience
functions wrapping the functions that return logical vectors in `which()`.

```{r}
locate_tag(sh_x, "[e]")

locate_any_tag(sh_x)

locate_no_tag(sh_x)
```
